---
title: "4 - Plots"
author: "JB Rickert"
date: "September 28, 2015"
output: html_document
---
## Base Graphics System
There are three plotting systems in R: 
1. Base R plots
2. Lattice / grid graphice plots
3. ggplot2

### Some very basic plots
Here are a few  basic plots:
    - a simple scatter plot
    - a histogram with normal curve
    - a boxplot 
    - kernel density plot
```{r}
par(mfrow=c(2,2))  						        # set up to draw multiple plots on the same chart
# Basic scatter plot
x <- rnorm(100)
y <- rnorm(100)
plot(x,y)

#Histogram with normal density superimposed
x <- rnorm(10000)							        # random draw from normal distribution
                                      # plot histogram
hist(x, freq = FALSE, col = "pink")			
curve(dnorm,								          # plot normal density
	  col = "dark blue",				        # set coor of curve
	  lwd=2,								            # fill in the area under the curve
	  add = TRUE)							          # add curve to existing plot
# 
# boxplot
boxplot(x,								  	        # draw a boxplot
	    col="yellow",						
		main="Box plot")						
# 
# Kernel density plot with rug
plot(density(x),							        # plot a kernel density estimate
	 main="Kernel Density Estimate")		# add a title
rug(x,col="red")  

```  

### Exploratory analysis with base graphics  

The Duncan data frame has 45 rows and 4 columns. Data on the prestige and other characteristics of 45 U. S. occupations in 1950.    

```{r}
library(car)  	# Load the library containing the sample data
dim(Duncan)			# How big is Duncan? 
head(Duncan)
attach(Duncan)  # Makes the columns available as variables in the glopal space.
pairs(cbind(prestige,income,education),  	# pairs is function that produces a maatrix of scatter plots
	panel=function(x,y){					          # define a function panel for the content of the matrix 
		points(x,y)							              # plot the points
		abline(lm(y~x), lty=2,col="blue")	    # add a linear regression 
		lines(lowess(x,y),col="red")		      # add a nonlinear regression
		},
	diag.panel=function(x){					        # define a new panel for the diagonals
		par(new=T)
		hist(x,main="",axes=F,nclass=12)	    # put a histogram on each diagonal
		}
)
```


## Lattice (trellis) Graphics

Lattice graphics are the second major plotting system in R. Plots built with lattice have a very distinctive look, but the real value is the ease of making trellis plots - graphs that display a variable conditioned on an other variable. Some useful websites are:
http://www.statmethods.net/advgraphs/trellis.html
http://user2007.org/program/presentations/sarkar.pdf   

### Lattice histograms

```{r}
library(lattice)
# Basic histogram
histogram( ~ prestige | type,  			# Formula to plot variable by a factor
	         data = Duncan,					  # Data set is an input
			 nint=10,
             layout = c(3,1), 
			 aspect = 1,
             xlab = "Type of Profession")

histogram( ~ income | type, 
	       data = Duncan,
		   nint=5,
           xlab = "Income",  
		   main= "Hitogram by profession type",
		   type = "density",
           panel = function(x, ...) {
              panel.histogram(x, ...)
              panel.mathdensity(dmath = dnorm, col = "black",
                                args = list(mean=mean(x),sd=sd(x)))
          } )
```  
### Lattice Level Plot   

```{r}
attach(environmental)
ozo.m <- loess((ozone^(1/3)) ~ wind * temperature * radiation,
       parametric = c("radiation", "wind"), span = 1, degree = 2)
w.marginal <- seq(min(wind), max(wind), length.out = 50)
t.marginal <- seq(min(temperature), max(temperature), length.out = 50)
r.marginal <- seq(min(radiation), max(radiation), length.out = 4)
wtr.marginal <- list(wind = w.marginal, temperature = t.marginal,
        radiation = r.marginal)
grid <- expand.grid(wtr.marginal)
grid[, "fit"] <- c(predict(ozo.m, grid))
contourplot(fit ~ wind * temperature | radiation, data = grid,
            cuts = 10, region = TRUE,
            xlab = "Wind Speed (mph)",
            ylab = "Temperature (F)",
            main = "Cube Root Ozone (cube root ppb)")
detach()
```   
### A 3D Lattice Plot

```{r}
## cloud.table
cloud(prop.table(Titanic, margin = 1:3),
      type = c("p", "h"), strip = strip.custom(strip.names = TRUE),
      scales = list(arrows = FALSE, distance = 2), panel.aspect = 0.7,
      zlab = "Proportion")[, 1]
```   

## ggplot2 Graphics

ggplot is the third major plotting system for R.
It is based on Leland Wilkinson's grammar of graphics. 
Plots are built in layers.

Some useful websites are:
http://www.cs.uic.edu/~wilkinson/TheGrammarOfGraphics/GOG.html
http://ggplot2.org/
http://docs.ggplot2.org/current/
http://www.cookbook-r.com/Graphs/

Here we illustrate building up a multi-layer plot
```{r}
library(ggplot2)
# Scatter plot
# Form the pase of the plot
p <- ggplot(Duncan, aes(income, prestige))  					
# p		# This will produce an error since no layers are defined									
# First layer specifies kind of plot
layer1 <- geom_point(colour="red")					
p + layer1		
# Second layer builds facet plot
layer2 <- facet_wrap(~ type)									        
p + layer1 + layer2
# Third layer add a regression line
layer3 <- geom_smooth(aes(group=type),method="lm",size=1,se=F)	
# Fourth layer add a title
layer4 <- ggtitle("Duncan Prestige Data")						   
p + layer1 + layer2 + layer3 + layer4
```

Here are ggplot2 analogs of the lattice plots above

```{r}
p <- ggplot(Duncan, aes(x = prestige, fill=type))
p + geom_histogram(binwidth=20,aes(y = ..density..),position="dodge")

# joran on StackOverflow
# http://bit.ly/1KgTXAU
library(plyr)
grid <- with(Duncan, seq(min(income), max(income), length = length(income)))
normaldens <- ddply(Duncan, "type", function(df) {
  data.frame( 
    predicted = grid,
    density = dnorm(grid, mean(df$income), sd(df$income))
  )
})

ggplot(Duncan, aes(x = income,fill = type)) +
  geom_histogram(aes(y = ..density..),binwidth = 15) +
  facet_grid(~ type) +
  geom_line(data = normaldens, aes(x = predicted, y = density))+
  scale_x_continuous(breaks=c(0,20,40,60,80)) + 
  ggtitle("Histogram by Profession Type") +
  theme(plot.title = element_text(lineheight=.8, face="bold"))

```

Adding a loess curve to "smooth" data

```{r}
# More examples from Steven Buechler of Notre Dame
# http://ntrda.me/1KtHGYG
dsmall <- diamonds[sample(nrow(diamonds), 500), ]
ggplot(data = dsmall, aes(carat, price)) +
       geom_point() + geom_smooth(method="loess") + 
       ggtitle("Sample of Diamonds Data with Smoother")	
```
